A non-additive generalization of probability theory \\for quantum mechanics and beyond

### 2024-10-11 Articles/2024-NonAdditiveProbability/NonAdditiveProbability.tex
NonAdditiveProbability: another pass at explaining the core issue

Each structure is important, but the balance between them is even more important. The key is to understand, on physical grounds, where each problem lies and therefore use the appropriate tool. For example, one may focus too much on the convex/linear structure and may be tempted to close the space under infinite convex/linear combinations. Mathematically, it would be very convenient. Physically, it is unsound. It is the link to experimental verification that tells us which limits are allowed, that particles cannot have an infinite average position. Therefore, it is the topology that has to tells us which limits are allowed, not the convex structure. Conceptual clarity has to come first. The math follows. A full examination of these issues is beyond the scope of this article, and we refer to our larger, and future, work. However, there is one key issue that we need to discuss as, we found, it was the main conceptual obstacle to progress.

On the convex structure we can define \textbf{pure states} as the ones that cannot be decomposed into other ensembles. Mathematically, these are the extreme points of the space and all other points can be understood as convex combinations, probability distributions, over them. So, for example, the pure states for a six-sided die are the six possible outcomes, and all ensembles are simply convex combinations $\sum_{i+1}^{6} p_i = 1$. This is exactly how classical discrete spaces work. As physicists, we are somehow trained to think that that's how everything works. In classical mechanics the pure states are perfect preparations of position and momentum. In a classical field theory, they are perfect preparation of all fields at each point in space. Things start breaking down in quantum mechanics, where we do have pure states, but then we also want pure states of position and momentum, which kind of work if you are careful, though we typically aren't. And then there is quantum field theory.

If we look closely, this does not make sense on physical grounds. The fact that the math becomes harder is simply a consequence of trying to fit a square peg in a round hole.

Perfect preparation of classical particles, continuous quantities in quantum mechanics and field configurations in any field theory are not things we can actually achieve. Physically, all ensembles we can prepare have finite precision and finite entropy, and those preparations have neither. They are not ensembles and they are not part of the a convex structure. In the classical case, we can understand them as infinitesimal decompositions: convex combinations that have more and more terms. This is true, but we cannot do that in the quantum case. If we have a pure state, this cannot be decomposed. Therefore an eigenstate of position is not a limit of decomposition, a limit in the convex structure, but a limit of a sequence of pure states, a limit in the topology, and the topology will not even include that limit. We may be able to see a pure state, in some sense, as a superposition of eigenstates of position, but that structure is only available in quantum mechanics, and it does not have a clear physical correspondence.

The insight is that probability distributions for a quantity are not defined by mixing the best ensembles we can prepare (i.e. a convex combination of extreme points), but by assigning a probability over each experimentally verifiable (i.e. open) interval of the possible values of said quantity (its spectra). The two coincide over a discrete set of orthogonal pure states, which is the least physically interesting case. If our are to be physically well defined and true generalizations, they must work in the more complicated case. We not only have to be able to talk about probability in terms of mixtures of ensembles, but we must be able to recover the interval of possible values for physical properties (i.e. the spectra), the object we typically assign probability two.

>>>>

Each structure is important, but the balance between them is even more important. The key is to understand, on physical grounds, where each problem lies and therefore use the appropriate tool. For example, one may focus on the convex/linear structure and may be tempted to close the space under infinite convex/linear combinations. Mathematically, it would be very convenient. However, not all infinite mixtures correspond to physically meaningful systems, such as a particle with an infinite average position. It is the connection to experimental verification, the topology, that will tell us which limits are allowed and which are not. This type of conceptual clarity has to come first. The math follows. A full examination of these issues is beyond the scope of this article, and we refer to our larger, and future, work. However, there is one key issue that we need to discuss as, we found, it was the main conceptual obstacle to progress.

On the convex structure we can define \textbf{pure states} as the ones that cannot be decomposed into other ensembles. For example, the pure states for a six-sided die are the ensemble perfectly prepared respectively on the six possible outcomes. All ensembles are convex combinations $\sum_{i+1}^{6} p_i = 1$ of those pure states. Mathematically, the pure states are the extreme points of the space and all other points can be understood as convex combinations over them. This is exactly how classical discrete spaces work. Most physicists, including are older selves, think that that's how everything works. For example, in classical mechanics the pure states are perfect preparations of position and momentum, and ensembles are probability distributions. This does not work.

We cannot consistently prepare classical particles perfectly, therefore the points of phase space are not ensemble. The mathematical facts that delta Dirac are not proper functions, have support with empty interior, have minus infinite entropy, and so on, are simply a consequence of our conceptual problem. A bit better is to think them as the limit of infinitesimal decomposition: classical ensembles can be decomposed into more and more precise ones, and the points of are the limit of this decomposition. Probability distributions are defined on these limits. This works in classical mechanics, but in general.

In quantum mechanics we have pure states, and these are ensembles that cannot be further decomposed into finer ensembles. Therefore an eigenstate of position is not a limit of decomposition, a limit in the convex structure. It is the limit where we are preparing a pure state on a smaller and smaller region of space. It is the limit of a sequence of pure states, a limit in the topology. Physically, more than mathematically, it is a completely different thing, therefore it is not a good route to generalization.\footnote{The fact that pure states can be seen, in some sense, as superposition of eigenstates of position, makes the matter even more confusing.}

The insight is that probability distributions for a quantity are not defined by statistical mixtures of perfect preparations (i.e. a convex combination of extreme points), but by assigning a probability over each experimentally verifiable (i.e. open) interval of the possible values of said quantity (its spectra). The two coincide in classical discrete space, which is the least physically interesting case as it does not even allow for classical particle mechanics.
